---
title: "Applied Stats Module 0 Project- 2024 GAC Softball Hitting Statistics"
author: "Jami Watson"
date: "2025-02-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
df<-read.csv("https://raw.githubusercontent.com/nurfnick/Data_Sets_For_Stats/refs/heads/master/CuratedDataSets/24GAChittingstatsUpdate.csv", fileEncoding = "UTF-8")

head(df)
```

```{r}
View(df)
```
I found this data at https://static.greatamericanconference.com/custompages/SOFTBALL%20Stats/2024/lgplyrs.htm. For this project we will be focusing on the quantitative variables RBIs, games played (GP), and games started (GS), and the categorical variables team and starters (GP/GS).
```{r}
summary(df)
```
Summary statistics of quantitative variable RBI's
```{r}
summary(df$RBI)
```
Standard Deviation of RBI's
```{r}
sd(df$RBI, na.rm = TRUE)
```
```{r}
hist(df$RBI, 
     main = "Histogram of RBI", 
     xlab = "RBI", 
     ylab = "Frequency", 
     col = "blue", 
     border = "black")
```

```{r}
boxplot(df$RBI, 
        main = "Box Plot of RBI", 
        ylab = "Values", 
        col = "lightblue", 
        border = "black", 
        horizontal = FALSE)
```
```{r}
qqnorm(df$RBI, main = "QQ Plot of RBI")
qqline(df$RBI, col = "red", lwd = 2)  # Adds a reference line
```

From the above graphical displays, I found that there were multiple outliers within the data and the distribution was heavily skewed. This is most likely due to the data including all players in the GAC, including starters, non-starters, and non-hitters.

Frequency and relative frequency tables for categorical variable, Teams. These values should closely align with roster count of each team in the GAC in the 2024 season.
```{r}
freq_table <- table(df$Team)
print(freq_table)
```
```{r}
rel_freq_table <- prop.table(freq_table)
print(rel_freq_table)
```
Two way table comparing games played and games started to determine number of starters on each team. A player is considered a starter if they start in 75% of games played. To be clear, this does not guarantee all starters are hitters, but it narrows down our data.
```{r}
df$Starters<-df$GS/df$GP>0.75
```
```{r}
table(df$Team,df$Starters)
```
Null Hypothesis- True mean is equivalent to 15 RBIs

Alternative Hypothesis- True mean is not equivalent to 15 RBIs
```{r}
t.test(df$RBI,mu=15)
```
We received a low p-value so we reject the null hypothesis. The true mean of RBIs among all players is not 15. 

Null Hypothesis- True mean among starters is equivalent to 15 RBIs

Alternative Hypothesis- True mean among starters is not equivalent to 15 RBIs
```{r}
t.test(df[df$Starters,"RBI"],mu=15)
```
We received a low p-value so we reject the null hypothesis. The true mean of RBIs among starters is not 15. However, I must say my estimate was very close to the true mean.

Null Hypothesis- True difference in means of RBIs of starters and non-starters is 0

Alternative Hypothesis- True difference in means of RBIs of starters and non-starters is not 0
```{r}
t.test(df[df$Starters,"RBI"],df[!df$Starters,"RBI"])
```
We received a low p-value so we reject the null hypothesis. The true difference in means of RBIs of starters and non-starters is not 0.


In conclusion, there is a significant difference in the mean number of RBIs of starters and non-starters, which accounts for why our initial data distribution of RBIs among players was so skewed. The mean number of RBIs among starters was not 15, however, I was very close with my estimate as the true mean was 15.929688. I suppose next I will remove non-hitters from my data set and see what results I get then.


