---
title: "Non-Parametric HW"
author: "Jami Watson"
date: "2025-05-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



```{r}
head(ChickWeight)
```

Wilcoxson Ranked Sign Test
Chickens that do not put on weight early in life are most likely to die. Examine the hypothesis that diet 1 will not have an effect on early weight gain of the chicks. Use a non-parametric method for matched pairs.

Null hypothesis= Median difference between weight at t=4 and t=0 from diet 1 is 0.
Alternatve hypothesis= Median difference between weight at t=4 and t=0 from diet 1 is not 0.

```{r}
table(ChickWeight$Diet == 1)
```
We have 220 rows in the table for diet 1

```{r}
df1 <- ChickWeight[ChickWeight$Diet == 1 & ChickWeight$Time %in% c(0,4), ]
```
So df1 contains only the rows from the ChickWeight dataset where the diet is 1 and time is 0 or 4.
```{r}
head(df1)
```
```{r}
library(tidyr)

wide_data <- pivot_wider(df1,
                         id_cols = Chick,
                         names_from = Time,
                         values_from = weight,
                         names_prefix = "Time_")

wide_data <- na.omit(wide_data)

print(wide_data)

wilcox.test(wide_data$Time_4, wide_data$Time_0, paired = TRUE, alternative = "two.sided")
```
Our p-value is low, although not very low, but anyways, we can reject the null hypothesis! Also WOW THAT WAS COMPLICATED MY HEAD HURTS.