---
title: "Non-Parametric HW"
author: "Jami Watson"
date: "2025-05-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



```{r}
head(ChickWeight)
```

Wilcoxson Ranked Sign Test
Chickens that do not put on weight early in life are most likely to die. Examine the hypothesis that diet 1 will not have an effect on early weight gain of the chicks. Use a non-parametric method for matched pairs.

Null hypothesis= Median difference between weight at t=4 and t=0 from diet 1 is 0.
Alternatve hypothesis= Median difference between weight at t=4 and t=0 from diet 1 is not 0.

```{r}
table(ChickWeight$Diet == 1)
```
We have 220 rows in the table for diet 1

```{r}
df1 <- ChickWeight[ChickWeight$Diet == 1 & ChickWeight$Time %in% c(0,4), ]
```
So df1 contains only the rows from the ChickWeight dataset where the diet is 1 and time is 0 or 4.
```{r}
head(df1)
```
```{r}
library(tidyr)

wide_data <- pivot_wider(df1,
                         id_cols = Chick,
                         names_from = Time,
                         values_from = weight,
                         names_prefix = "Time_")

wide_data <- na.omit(wide_data)

print(wide_data)

wilcox.test(wide_data$Time_4, wide_data$Time_0, paired = TRUE, alternative = "two.sided")
```
Our p-value is low, although not very low, but anyways, we can reject the null hypothesis! So diet 1 does effect early weight gain in chicks. Also WOW THAT WAS COMPLICATED MY HEAD HURTS.

Now onto the kruskall-wallis test
Examine the claim that there is a difference in early weight gain between the different diets. Use a non-parametric method for comparing multiple medians. Be sure to include an appropriate visualization.

Null hypothesis= The median difference between weight at t=4 and t=0 for all of the diets is 0.
Alternative hypothesis= The median difference between weight at t=4 and t=0 for all of the diets is not 0.

```{r}
df3 <- ChickWeight[ChickWeight$Time %in% c(0,4), ]
head(df3)
```
```{r}
library(tidyr)
library(dplyr)

diet_info <- df3 %>% 
  filter(Time == 0) %>% 
  select(Chick, Diet)

wide_data <- df3 %>%
  pivot_wider(
    id_cols = Chick,
    names_from = Time,
    values_from = weight,
    names_prefix = "Time_"
  )
wide_data <- left_join(wide_data, diet_info, by = "Chick")

wide_data$weight_change <- wide_data$Time_4 - wide_data$Time_0

print(wide_data)
```
```{r}
boxplot(wide_data$weight_change ~ wide_data$Diet)
```
```{r}
by(wide_data$weight_change,wide_data$Diet,median)
```
We get NA here for diet 1, this is because chick 18 which was on diet 1, only has values for weight at time 0 and 2. Not a big deal, diet 1 obviously had the least early chick weight gain anyways. However, this may effect our test and we may have to adjust to times 0 and 2 to get a clear look.
```{r}
kruskal.test(weight_change ~ Diet , data = wide_data)
```
We get a very low p-value so we reject the null hypothesis. The median difference between weight at t=4 and t=0 for all of the diets is not 0. Diet has an effect on early weight gain in chicks. This test was much less of a head ache.